Lack of Evidence?: Academic Probation Does Not Affect Student Dropout and Subsequent Performance

### 2021-04-24 outputs/paper/paper.Rmd
Commit for Discussion

---
title: "Lack of Evidence?: Academic Probation Does Not Affect Student Dropout and Subsequent Performance"
subtitle: "Revisiting the Ability, Gender, and Performance Standards: Evidence from Academic Probation (Lindo, Sanders, and Oreopoulos 2010)"
author: "Hong Shi"
date: "`r format(Sys.time(), '%d %B %Y')`"
abstract: "Many universities use academic probation as a wake-up call to ensure currently enrolled students to achieve minimum academic standards. This paper examines the causal impact on first-year students' responses in a Canadian university after being placed on academic probation in a sharp regression discontinuity design. With adjusted bandwidth, the results do not suggest that placing first year students on academic probation affects the dropout decision, subsequent annual GPA and graduation. While there are heterogeneous responses across students' highschool performance, gender, first language towards academic probation. #High Level Importance \\par\\textbf{Keywords:} Regression Discontinuity Design, Bandwidth, Academic Probation, Student Performance, P-value, Causality"
output:
  bookdown::pdf_document2:
    latex_engine: xelatex
toc: FALSE

@@

&nbsp; &nbsp; The rest of this paper is organized as follows: In Data section (Section \@ref(Data)), I would first explain the domain background about the academic probation. Next, I introduce the data used in the original paper. And then, I would describe the Quasi-experiment design rationale and also raise concerns about the bandwidth used in the original paper. In Model section (Section \@ref(mainmodel)), I elaborate on the sharp RDD, address its assumptions and propose the model used in this study. In Results section (Section \@ref(result)), I show the impact of academic probation on student dropout and subsequent academic performance with the adjusted bandwidth. In Discussion section (Section \@ref(discuss)), I would first comment on the study results, and then elaborate on the causality of this quasi-experiment, address internal and external validity regarding the campus variation. I would also discuss potential bias in this experiment. Finally, I conclude and suggest future directions of this study. ^[Codes and data are available at the GitHub repo: https://github.com/honn-ishinn/RDD_academic_probation.]

@@

&nbsp; &nbsp; The data comes from a large Canadian university with three individual campuses: one central campus (Campus 1) and two smaller satellite campuses (Campus 2 and Campus 3) from 1996 to 2005. I analyzed it using R [@citeR], and packages `tidyverse` [@cite_tidyverse],  `here` [@cite_here], `rdrobust` [@cite_rdrobust], `rdd`[@cite_rdd], `rdpower`[@cite_rdpower], `haven`[@cite_haven] . I used packages `bookdown` [@cite_bookdown], `kableExtra` [@cite_kableextra], , `finalfit` [@cite_finalfit], `modelsummary`[@cite_modelsummary], `broom` [@cite_broom], `ggpubr` [@cite_ggpubr] to format the document and referenced `Impact Evaluation in Practice` [@cite_impact] to evaluate this experiment.
```{r, include=FALSE}
clean_data <- read.csv(here("inputs/data/cleaned_original.csv"))
```

@@

&nbsp; &nbsp; Students on academic probation will be suspended in subsequent sessions if their grades do not improve. At all campuses, students on probation can avoid suspension and return to good academic standing by improving their cumulative GPA up to the cutoff. Students who fail to sufficiently improve their grades are suspended for one full academic year. If suspended students return to the university and again fail to sufficiently improve their grades, they can be suspended for three years. A third failure to meet the GPA requirement can lead to permanent suspension from all campuses. Students will be notified after being placed on academic probation. A sample letter sent to students at Campus 2 is in Appendix (Appendix 1 \@ref(letter)).

@@

```{r allsummary, tab.cap="Summary Statistics of Observable Characteristics and Outcomes of All Students", echo = FALSE, message = FALSE, warning=FALSE}
# Generate summary table
kable(combine1[,c(3,1,2)],caption = "Summary Statistics of Observable Characteristics and Outcomes of All Students",
       booktabs = TRUE, linesep = "") %>% 
  kable_styling() %>% 
  add_footnote("For all characteristics and outcomes except graduation rates and distance from cutoff in 2nd year. The entire dataset consists of 44362 students. Graduation rate samples are 30017 for Year 4, 24581 for Year 5, 19757 for Year 6.38576 students have GPA observed in 2nd Year", notation = "number")
```

&nbsp; &nbsp; After the data cleaning , the original data includes a total of 44362 observations and Table \@ref(tab:allsummary) shows the summary statistics of student observable characteristics and outcomes such as gender, age, birth place, study campus, first language, attempted course credit^[This Canadian university requires at least 20 course credits to graduate.], registration status, GPA, academic standing, high school performance, and graduation status. Since the probation cutoff among campus is different (1.5 in Campus 1 and 2, 1.6 in Campus 3), the data also includes the variable of distance from the campus probation cutoff. Instead of the absolute GPA, the distance from cutoff variable will be used as the forcing variable in the regression discontinuity design which will be discussed in the following section.

@@

&nbsp; &nbsp; In this study, since the RDD only provides the most accurate estimates around the probation cutoff where treatment and comparison are most similar [@cite_impact], it is critical to determine the bandwidth(i.e., students within certain grade points of the probation cutoff score) to maintain the balance of observed characteristics of students around the probation cutoff. In Lindo's original paper, they choose the bandwidth of 0.6 and the following table (Table \@ref(tab:lindo)) shows the summary statistics of student observable characteristics within 0.6 GPA.

@@

&nbsp; &nbsp; The following table (Table \@ref(tab:gradingtheme)) shows the grading theme of this Canadian university^[This is the current grading theme of this university as of 2021. The grading theme between 1996 and 2005 might be different and could not be retrieved on the university website. While the current grading theme may still serve as a reference of this study]. By applying Lindo's bandwidth of 0.6 GPA, observations barely above the probation cutoff roughly covers students within 60-66 average grade percentage (1.5-2.1 GPA in Campus 1 and 2, 1.6-2.2 GPA in Campus 3), and observations barely below the probation cutoff roughly covers students within 53-59 average grade percentage (0.9-1.5 GPA in Campus 1 and 2, 1.0-1.6 GPA in Campus 3). By converting the grade point value into grade percentage, it might become more obvious that there exists academic performance difference between students in good standing with 66 average grade percentage and students on probation with 53 average grade percentage. So I would argue that the bandwidth used in Lindo's is too wide to maintain similar characteristics, particularly academic potentials, between the probation and good standing students.

```{r gradingtheme, fig.cap='Grading Scheme of the University', echo = FALSE}
# Grading scheme obtained from the university website
tibble(
  "Grade" = list("A+", "A", "A-", "B+", "B", "B-", "C+", "C", "C-", "D+", "D", "D-", "F"),
  "Grade Point Value" = list("4.0", "4.0", "3.7", "3.3", "3.0", "2.7", "2.3", "2.0", "1.7", "1.3", "1.0", "0.7", "0.0"),
  "Grade Percentage" = list("90-100", "85-89","80-84","77-79", "73-76","70-72","67-69","63-66","60-62","57-59","53-56","50-52","0-49"),
  "Definition" = list("Excellent","","","Good","","","Adequate","","","Marginal","","","Inadequate; no credit obtained")
) %>% 
  kable(caption = "Grading Scheme of the University",
        align = "cccc") %>% 
  kable_styling(latex_options = "hold_position")
```

@@

1. Balance the observed characteristics between good standing and probation students within the bandwidth.

@@

```{r mycutoff, tab.cap="Summary Statistics of Students Within 0.2 GPA of Probation Cutoff", echo = FALSE, message = FALSE, warning=FALSE}
kable(combine2[,c(4,1,2,3)],caption = "Summary Statistics of Students Within 0.2 GPA of Probation Cutoff",
       booktabs = TRUE, linesep = "") %>% 
  kable_styling(latex_options = "hold_position")
```

&nbsp; &nbsp; Since the probation cutoff GPA lies between "adequate" C- of 1.7 GPA and "marginal" D+ of 1.3 GPA, I adjust the bandwidth to 0.2 GPA to reduce the academic performance difference. Table \@ref(tab:mycutoff) shows the summary statistics of 4106 student observable characteristics within 0.2 GPA. Even though the t-test (Table \@ref(tab:ttesthsme)) on highschool grade percentile still suggests difference on highschool performance, the estimated highschool grade percentile difference is much less than those within 0.6 bandwidth. Many probation students with low highschool grade are excluded after applying the narrower 0.2 GPA bandwidth as shown in the following histogram (Figure \@ref(fig:combinehs)). Besides, the t-test (Table \@ref(tab:ttestcrme)) on course credit does not show a significant difference in course load within 0.2 bandwidth, so the observed characteristics become more balanced.

@@

```{r, include=FALSE}
# Set students on academic probation as the baseline group for easier visualizations
clean_cut$status <- as.factor(clean_cut$status)
clean_cut$status <- relevel(clean_cut$status, "On Academic Probation")
clean_cut2$status <- as.factor(clean_cut2$status)
clean_cut2$status <- relevel(clean_cut2$status, "On Academic Probation")
```

```{r combinehs, fig.align='center', fig.cap="Distribution of Highschool Grade Within Certain GPA Bandwidth", fig.width=10, fig.height=5, echo= FALSE, message=FALSE, warning=FALSE}
# Referenced the following  http://www.sthda.com/english/articles/32-r-graphics-essentials/126-combine-multiple-ggplots-in-one-graph/
# combine figure into 1
ggarrange(hs_06,hs_02, ncol = 2, nrow = 1, common.legend = TRUE, legend = "bottom")
```

&nbsp; &nbsp; The power of a test is the probability that the test will reject the test null hypothesis to detect a correct alternative hypothesis [@cite_stats_tests]. In this study, as the length of the bandwidth increases, more students are included within the bandwidth. The increase in the number of student raises the power to detect the impact on placing students on probation so that the impact estimation is less likely to happen by chance rather than by the probation cutoff rule. Therefore, I use the `rdpower`[@cite_rdpower] package to calculate the sufficient sample size required (i.e., effect size) to detect the probation effect on (1) Student dropout, (2) Second year GPA and (3) Graduation rate by Year 4. The result as shown below (Table \@ref(tab:powersample)) suggests that the bandwidth of 0.2 GPA with 4106 students (1745 on probation, 2361 in good standing) could raise sufficient statistical power. Further adjustment on the bandwidth might either unbalance student characteristics (Increase bandwidth) or decrease the power of the study (Decrease bandwidth). So I would apply the bandwidth of 0.2 GPA to examine the academic probation impact in the following sessions.

@@

3. The forcing variable should be continuous [@cite_Rohan_Chapter15]: as mentioned in Data Section \@ref(od), this study use student GPA distance from cutoff as the forcing variable. Since the GPA is a continuous variable, the GPA distance from cutoff is also continuous so this assumption is satisfied.

@@

&nbsp; &nbsp; Since the academic probation cutoff rule is known, one may question that students barely below the probation cutoff would study harder to avoid being placed on academic probation. However, the survey conducted in Lindo's paper suggests that first year students are generally less familiar with university policies to know exact GPA required to avoid academic probation. And Lindo also finds no evidence that students' grades are related to their knowledge about academic probation [@cite_original]. So first year students are not likely to "play with" the university probation policy to avoid academic probation. The following figure (Figure \@ref(fig:histcutoff)) shows the distribution of the GPA distance from probation cutoff of all students. There is a smooth increase on the frequency distribution of the student GPA until reaching students' average GPA (GPA distance from probation around 1) and there is no obvious jump on either side of the probation cutoff. So students on either side of the probation cutoff is balanced as suggested in Data Section \@ref(bandwidth).

@@

$$Y_{i} = \beta_{0}\mbox{Proation}_{i} + f(\mbox{Cutoff}\:\mbox{Distance}_{i}) + \alpha_{1}\mbox{Gender}_{i}+\alpha_{2}\mbox{HighSchool}_{i}+\alpha_{3}\mbox{Language}_{i} +\epsilon_{i}$$

where

@@

* $\mbox{Gender}$ is an indicator variable to show whether the student $i$ is male or female. Its coefficient $\alpha_{1}$ is the estimate of gender difference in response to academic probation.

@@

&nbsp; &nbsp; As the GPA is an continuous variable from 0.0 to 4.0, I would use the linear regression to estimate the probation impact on student annual GPA in the second academic year.

@@

&nbsp; &nbsp; As mentioned in Model Section \@ref(model), logistic regression is used to estimate the probation impact on probability of student dropout.
$$Prob(\mbox{Dropout}_{i}) = \beta_{0}\mbox{Proation}_{i} + f(\mbox{Cutoff}\:\mbox{Distance}_{i}) + \alpha_{1}\mbox{Gender}_{i}+\alpha_{2}\mbox{HighSchool}_{i}+\alpha_{3}\mbox{Language}_{i} +\epsilon_{i}$$

@@

```{r dropfig, fig.align='center', fig.cap="Academic Probation Impact on Dropout", fig.width=8, fig.height=4, echo= FALSE, message=FALSE, warning=FALSE}
ggplot(data= clean_cut, aes(x = dist_from_cut, y = left_school, color = status))+
  geom_point(alpha = 0.05)+
  geom_smooth(method = "glm", method.args = list(family = 'binomial'))+
  theme_light()+
  geom_vline(xintercept = 0, linetype = "dotted")+
  theme(legend.title = element_blank(), legend.position = "bottom", plot.title = element_text(hjust = 0.5))+
  guides(color=guide_legend(override.aes=list(fill=NA)))+
  labs(title = "Academic Probation Impact on Dropout",
       x = "Year 1 GPA minus Probation Cutoff",
       y = "Left University")
```

@@

&nbsp; &nbsp; As mentioned in Model Section \@ref(model), linear regression is used to estimate the probation impact on student second year annual GPA.
$$\mbox{GPAYear2}_{i} = \beta_{0}\mbox{Proation}_{i} + f(\mbox{Cutoff}\:\mbox{Distance}_{i}) + \alpha_{1}\mbox{Gender}_{i}+\alpha_{2}\mbox{HighSchool}_{i}+\alpha_{3}\mbox{Language}_{i} +\epsilon_{i}$$

@@

&nbsp; &nbsp; Within 0.2 GPA bandwidth, there is significant difference on student second year GPA across gender and highschool performance. Compared to female students, male students earn 0.120 less annual GPA in the second year, holding other baseline characteristics constant. Compared to students with relative poor highschool academic performance, students with satisfactory highschool academic performance earn 0.099 more annual GPA in the second year, holding other baseline characteristics constant. No other significant difference is found on student first language.
```{r, include=FALSE}
# Linear Regression to estimate the impact of probation on student Year 2 GPA

@@

&nbsp; &nbsp; As mentioned in Model Section \@ref(model), logistic regression is used to estimate the probation impact on probability of student graduating by Year 4.
$$Prob(\mbox{Graduation}_{i}) = \beta_{0}\mbox{Proation}_{i} + f(\mbox{Cutoff}\:\mbox{Distance}_{i}) + \alpha_{1}\mbox{Gender}_{i}+\alpha_{2}\mbox{HighSchool}_{i}+\alpha_{3}\mbox{Language}_{i} +\epsilon_{i}$$
&nbsp; &nbsp; The model baseline characteristics are the same as mentioned in dropout Section \@ref(baseline). As shown in the following table (Table \@ref(tab:gradsummary)), within 0.2 GPA bandwidth, compared to students on academic probation, students in good standing have $26.8\%$ increase in the odds of graduating by Year 4, while holding other baseline characteristics constant. **However**, the p-value of 0.161 suggests that the graduation does not significantly differ between good standing students and probation students, holding other baseline characteristics constant. And this difference could also not be clearly detected around the probation cutoff as shown(Figure \@ref(fig:gradfig)). Within 0.3 GPA and 0.6 GPA bandwidth, the results also do not suggest difference on graduation between good standing students and probation students at 0.05 significance level, holding other baseline characteristics constant.

@@

```{r gradfig, fig.align='center', fig.cap="Academic Probation Impact on Graduation by Year 4", fig.width=6, fig.height=3, echo= FALSE, message=FALSE, warning=FALSE}
clean_cut$probation_year1 <- as.factor(clean_cut$probation_year1)
ggplot(data= clean_cut, aes(x = dist_from_cut, y = gradin4, color = status))+
  geom_point(alpha = 0.05)+
  geom_smooth(method = "glm", method.args = list(family = 'binomial'))+
  theme_light()+
  geom_vline(xintercept = 0, linetype = "dotted")+
  theme(legend.title = element_blank(), legend.position = "bottom", plot.title = element_text(hjust = 0.5))+
  guides(color=guide_legend(override.aes=list(fill=NA)))+
  labs(title = "Academic Probation Impact on Graduation by Year 4",
       x = "Year 1 GPA minus Probation Cutoff",
       y = "Graduate by Year 4")
```

@@

&nbsp; &nbsp; Since all of the result interpretations are based on p-values, reassessing the definition of p-value may help to explain the above inconsistent results. In simple common English, P-value means the probability that something checked wrong to happen. The lower p-value means a higher probability of something checked right to be true, and "is there an academic probation impact on first year student outcome" is the checkpoint in this study.

@@

&nbsp; &nbsp; According to the student summary statistics (Table \@ref(tab:lindo)) and t-test conducted on students highschool performance (Table \@ref(tab:ttesthslin)) and credit attempted in the first year (Table \@ref(tab:ttestcrlin)) in Data Section \@ref(bandwidth), within the bandwidth of 0.6 GPA, students below the cutoff generally hold weaker academic potentials than those above the cutoff. And these academic differences also serves as key factors affecting student outcomes.

&nbsp; &nbsp; Even though Lindos's original paper did test whether students share similar observable characteristics around the probation cutoff GPA in Results Section A [@cite_original], it failed to access students characteristics that are further away from the cutoff. It is unlikely that a student with 1.0 GPA is as similar as one with 2.0 GPA, while they are all included within Lindo's bandwidth to estimate the impact of academic probation. As a result, using a wider bandwidth could violate the similarity assumption of RDD, and interpreting p-values from assumption violated models to seek "statistical significance" may become senseless.

@@

&nbsp; &nbsp; The Credit/No Credit option allows students to remove course result from their GPA but still count course credit if passing the course(at least $50\%$ final mark, 2.0 credits maximum) before a certain deadline (usually in the middle of the semester). For example, if a student utilizes this policy on:

@@

&nbsp; &nbsp; Above examples demonstrate a relatively poor quality of data for analysis. And these dirty data may also affect accuracy of study results.

>>>>

---
title: "Lack of Evidence?: Academic Probation Does Not Affect Student Dropout and Subsequent Performance"
subtitle: "Revisiting the Ability, Gender, and Performance Standards: Evidence from Academic Probation (Lindo, Sanders, and Oreopoulos 2010)"
author: "Hong Shi"
date: "April 24 2021"
abstract: "Many universities use academic probation as a wake-up call to ensure currently enrolled students to achieve minimum academic standards. This paper examines the causal impact on first-year students' responses in a Canadian university after being placed on academic probation in a sharp regression discontinuity design. With adjusted bandwidth, the results do not suggest that placing first year students on academic probation affects the dropout decision, subsequent annual GPA and graduation. While there are heterogeneous responses across students' highschool performance, gender, first language towards academic probation. \\par\\textbf{Keywords:} Regression Discontinuity Design, Bandwidth, Academic Probation, Student Performance, P-value, Causality"
output:
  bookdown::pdf_document2:
    latex_engine: xelatex
toc: FALSE

@@

&nbsp; &nbsp; The rest of this paper is organized as follows: In Data section (Section \@ref(Data)), I would first explain the domain background about the academic probation. Next, I introduce the data used in the original paper. And then, I would describe the Quasi-experiment design rationale and raise concerns about the bandwidth used in the original paper. In Model section (Section \@ref(mainmodel)), I elaborate on the sharp RDD, address its assumptions and propose the model used in this study. In Results section (Section \@ref(result)), I show the impact of academic probation on student dropout and subsequent academic performance with the adjusted bandwidth. In Discussion section (Section \@ref(discuss)), I would first comment on the study results, and then elaborate on the causality of this quasi-experiment, address internal and external validity regarding the campus variation. I would also discuss potential bias in this experiment. Finally, I conclude and suggest future directions of this study. ^[Codes and data are available at the GitHub repo: https://github.com/honn-ishinn/RDD_academic_probation.]

@@

&nbsp; &nbsp; The data comes from a large Canadian university with three individual campuses: one central campus (Campus 1) and two smaller satellite campuses (Campus 2 and Campus 3) from 1996 to 2005. I analyzed it using R [@citeR], and packages `tidyverse` [@cite_tidyverse],  `here` [@cite_here], `rdrobust` [@cite_rdrobust], `rdd` [@cite_rdd], `rdpower` [@cite_rdpower], `haven` [@cite_haven] . I used packages `bookdown` [@cite_bookdown], `kableExtra` [@cite_kableextra], `finalfit` [@cite_finalfit], `modelsummary` [@cite_modelsummary], `broom` [@cite_broom], `ggpubr` [@cite_ggpubr] to format the document and referenced `Impact Evaluation in Practice` [@cite_impact] to evaluate this experiment. I used `Shiny` [@cite_shiny] for enhancement to display interactive model results (Discussion Section \@ref(enhancement)).
```{r, include=FALSE}
clean_data <- read.csv(here("inputs/data/cleaned_original.csv"))
```

@@

&nbsp; &nbsp; Students on academic probation will be suspended in subsequent sessions if their grades do not improve. At all campuses, students on probation can avoid suspension and return to good academic standing by improving their cumulative GPA up to the cutoff. Students who fail to sufficiently improve their grades are suspended for one full academic year. If suspended students return to the university and again fail to sufficiently improve their grades, they can be suspended for three years. A third failure to meet the GPA requirement can lead to permanent suspension from all campuses. Students will be notified after being placed on academic probation. A sample letter sent to students at Campus 2 is in Appendix (Appendix \@ref(letter)).

@@

```{r allsummary, tab.cap="Summary Statistics of Observable Characteristics and Outcomes of All Students", echo = FALSE, message = FALSE, warning=FALSE}
# Generate summary table
kable(combine1[,c(3,1,2)],caption = "Summary Statistics of Observable Characteristics and Outcomes of All Students",
       booktabs = TRUE, linesep = "") %>% 
  kable_styling() %>% 
  add_footnote("For all characteristics and outcomes except graduation rates and distance from cutoff in 2nd year. The entire dataset consists of 44362 students. Graduation rate samples are 30017 for Year 4, 24581 for Year 5, 19757 for Year 6.38576 students have GPA observed in 2nd Year", notation = "number")
```
\newpage

&nbsp; &nbsp; After the data cleaning, the original data includes a total of 44362 observations and Table \@ref(tab:allsummary) shows the summary statistics of student observable characteristics and outcomes such as gender, age, birth place, study campus, first language, attempted course credit^[This Canadian university requires at least 20 course credits to graduate.], registration status, GPA, academic standing, high school performance, and graduation status. Since the probation cutoff among campus is different (1.5 GPA in Campus 1 and 2, 1.6 GPA in Campus 3), the data also includes the variable of distance from the campus probation cutoff. Instead of the absolute GPA, the distance from cutoff variable will be used as the forcing variable in the regression discontinuity design which will be discussed in the following section.

@@

&nbsp; &nbsp; In this study, since the RDD only provides the most accurate estimates around the probation cutoff where treatment and comparison are most similar [@cite_impact], it is critical to determine the bandwidth(i.e., students within certain grade points of the probation cutoff score) to maintain the balance of observed characteristics of students around the probation cutoff. In Lindo's original paper, they choose the bandwidth of 0.6 GPA and the following table (Table \@ref(tab:lindo)) shows the summary statistics of student observable characteristics within 0.6 GPA.

@@

&nbsp; &nbsp; The following table (Table \@ref(tab:gradingtheme)) shows the grading theme of this Canadian university^[This is the current grading theme of this university as of 2021. The grading theme between 1996 and 2005 might be different and could not be retrieved on the university website. While the current grading theme may still serve as a reference of this study]. By applying Lindo's bandwidth of 0.6 GPA, observations barely above the probation cutoff roughly covers students within 60-66 average grade percentage (1.5-2.1 GPA in Campus 1 and 2, 1.6-2.2 GPA in Campus 3), and observations barely below the probation cutoff roughly covers students within 53-59 average grade percentage (0.9-1.5 GPA in Campus 1 and 2, 1.0-1.6 GPA in Campus 3). By converting grade point value into grade percentage, it might become more obvious that there exists academic performance difference between students in good standing with 66 average grade percentage and students on probation with 53 average grade percentage. Therefore, I would argue that the bandwidth used in Lindo's is too wide to maintain similar characteristics, particularly academic potentials, between the probation and good standing students.

```{r gradingtheme, fig.cap='Grading Scheme of the University', echo = FALSE}
# Grading scheme obtained from the university website
tibble(
  "Grade" = list("A+", "A", "A-", "B+", "B", "B-", "C+", "C", "C-", "D+", "D", "D-", "F"),
  "Grade Point Value" = list("4.0", "4.0", "3.7", "3.3", "3.0", "2.7", "2.3", "2.0", "1.7", "1.3", "1.0", "0.7", "0.0"),
  "Grade Percentage" = list("90-100", "85-89","80-84","77-79", "73-76","70-72","67-69","63-66","60-62","57-59","53-56","50-52","0-49"),
  "Definition" = list("Excellent","","","Good","","","Adequate","","","Marginal","","","Inadequate; no credit obtained")
) %>% 
  kable(caption = "Grading Scheme of the University",
        align = "cccc") %>% 
  kable_styling(latex_options = "hold_position")
```
\newpage

@@

1. Balance characteristics between good standing and probation students within the bandwidth.

@@

```{r mycutoff, tab.cap="Summary Statistics of Students Within 0.2 GPA of Probation Cutoff", echo = FALSE, message = FALSE, warning=FALSE}
kable(combine2[,c(4,1,2,3)],caption = "Summary Statistics of Students Within 0.2 GPA of Probation Cutoff",
       booktabs = TRUE, linesep = "") %>% 
  kable_styling(latex_options = "hold_position")
```
```{r, include=FALSE}
# Set students on academic probation as the baseline group for easier visualizations
clean_cut$status <- as.factor(clean_cut$status)
clean_cut$status <- relevel(clean_cut$status, "On Academic Probation")
clean_cut2$status <- as.factor(clean_cut2$status)
clean_cut2$status <- relevel(clean_cut2$status, "On Academic Probation")
```

&nbsp; &nbsp;
&nbsp; &nbsp;
```{r combinehs, fig.align='center', fig.cap="Distribution of Highschool Grade Within Certain GPA Bandwidth", fig.width=10, fig.height=5, echo= FALSE, message=FALSE, warning=FALSE}
# Referenced the following  http://www.sthda.com/english/articles/32-r-graphics-essentials/126-combine-multiple-ggplots-in-one-graph/
# combine figure into 1
ggarrange(hs_06,hs_02, ncol = 2, nrow = 1, common.legend = TRUE, legend = "bottom")
```

&nbsp; &nbsp; Since the probation cutoff GPA lies between "adequate" C- of 1.7 GPA and "marginal" D+ of 1.3 GPA, I adjust the bandwidth to 0.2 GPA to reduce the academic performance difference. Table \@ref(tab:mycutoff) shows the summary statistics of 4106 student observable characteristics within 0.2 GPA. Even though the t-test (Table \@ref(tab:ttesthsme)) on highschool grade percentile still suggests difference on highschool performance, the estimated highschool grade percentile difference is much less than those within 0.6 bandwidth. Many probation students with low highschool grade are excluded after applying the narrower 0.2 GPA bandwidth as shown in the above histogram (Figure \@ref(fig:combinehs)). Besides, the t-test (Table \@ref(tab:ttestcrme)) on course credit does not show a significant difference in course load within 0.2 bandwidth, so the observed characteristics become more balanced.

@@

&nbsp; &nbsp; The power of a test is the probability that the test will reject the test null hypothesis to detect a correct alternative hypothesis [@cite_stats_tests]. In this study, as the length of the bandwidth increases, more students are included within the bandwidth. The increase in the number of students raises the power to detect the impact on placing students on probation so that the impact estimation is less likely to happen by chance rather than by the probation cutoff rule. Therefore, I use the `rdpower`[@cite_rdpower] package to calculate the sufficient sample size required (i.e., effect size) to detect the probation effect on (1) Student dropout, (2) Second year GPA and (3) Graduation rate by Year 4. The result as shown below (Table \@ref(tab:powersample)) suggests that the bandwidth of 0.2 GPA with 4106 students (1745 on probation, 2361 in good standing) could raise sufficient statistical power. Further adjustment on the bandwidth might either unbalance student characteristics (Increase bandwidth) or decrease the power of the study (Decrease bandwidth). So I would apply the bandwidth of 0.2 GPA to examine the academic probation impact in the following sessions.

@@

3. The forcing variable should be continuous [@cite_Rohan_Chapter15]: as mentioned in Data Section \@ref(od), this study uses student GPA distance from cutoff as the forcing variable. Since the GPA is a continuous variable, the GPA distance from cutoff is also continuous so this assumption is satisfied.

@@

&nbsp; &nbsp; Since the academic probation cutoff rule is known, one may question that students barely below the probation cutoff would study harder to avoid being placed on academic probation. However, the survey conducted in Lindo's paper suggests that first year students are generally less familiar with university policies to know exact GPA required to avoid academic probation. And Lindo also finds no evidence that students' grades are related to their knowledge about academic probation [@cite_original]. So first year students are NOT likely to "play with" the university probation policy to avoid academic probation. The following figure (Figure \@ref(fig:histcutoff)) shows the distribution of the GPA distance from probation cutoff of all students. There is a smooth increase on the frequency distribution of the student GPA until reaching students' average GPA (GPA distance from probation around 1) and there is no obvious jump on either side of the probation cutoff. So students on either side of the probation cutoff is balanced as suggested in Data Section \@ref(bandwidth).

@@

$$Y_{i} = \beta_{0}\mbox{Probation}_{i} + f(\mbox{Cutoff}\:\mbox{Distance}_{i}) + \alpha_{1}\mbox{Gender}_{i}+\alpha_{2}\mbox{HighSchool}_{i}+\alpha_{3}\mbox{Language}_{i} +\epsilon_{i}$$
where

@@

* $\mbox{Gender}_{i}$ is an indicator variable to show whether the student $i$ is male or female. Its coefficient $\alpha_{1}$ is the estimate of gender difference in response to academic probation.

@@

&nbsp; &nbsp; As the GPA is a continuous variable from 0.0 to 4.0, I would use the linear regression to estimate the probation impact on student annual GPA in the second academic year.

@@

&nbsp; &nbsp; As mentioned in Model Section \@ref(model), logistic regression is used to estimate the probation impact on probability of student dropout.
$$Prob(\mbox{Dropout}_{i}) = \beta_{0}\mbox{Probation}_{i} + f(\mbox{Cutoff}\:\mbox{Distance}_{i}) + \alpha_{1}\mbox{Gender}_{i}+\alpha_{2}\mbox{HighSchool}_{i}+\alpha_{3}\mbox{Language}_{i} +\epsilon_{i}$$

@@

&nbsp; &nbsp;
&nbsp; &nbsp;

```{r dropfig, fig.align='center', fig.cap="Academic Probation Impact on Dropout", fig.width=8, fig.height=4, echo= FALSE, message=FALSE, warning=FALSE}
ggplot(data= clean_cut, aes(x = dist_from_cut, y = left_school, color = status))+
  geom_point(alpha = 0.05)+
  geom_smooth(method = "glm", method.args = list(family = 'binomial'))+
  theme_light()+
  geom_vline(xintercept = 0, linetype = "dotted")+
  theme(legend.title = element_blank(), legend.position = "bottom", plot.title = element_text(hjust = 0.5))+
  guides(color=guide_legend(override.aes=list(fill=NA)))+
  labs(title = "Academic Probation Impact on Dropout",
       x = "Year 1 GPA minus Probation Cutoff",
       y = "Left University")
```
\newpage

@@

&nbsp; &nbsp; As mentioned in Model Section \@ref(model), linear regression is used to estimate the probation impact on student second year annual GPA.
$$\mbox{GPAYear2}_{i} = \beta_{0}\mbox{Probation}_{i} + f(\mbox{Cutoff}\:\mbox{Distance}_{i}) + \alpha_{1}\mbox{Gender}_{i}+\alpha_{2}\mbox{HighSchool}_{i}+\alpha_{3}\mbox{Language}_{i} +\epsilon_{i}$$

@@

&nbsp; &nbsp; Within 0.2 GPA bandwidth, there is significant difference on student second year GPA across gender and highschool performance. Compared to female students, male students earn 0.120 less annual GPA in the second year, holding other baseline characteristics constant. Compared to students with relatively poor highschool academic performance, students with satisfactory highschool academic performance earn 0.099 more annual GPA in the second year, holding other baseline characteristics constant. No other significant difference is found on student first language.
```{r, include=FALSE}
# Linear Regression to estimate the impact of probation on student Year 2 GPA

@@

&nbsp; &nbsp; As mentioned in Model Section \@ref(model), logistic regression is used to estimate the probation impact on probability of student graduating by Year 4.
$$Prob(\mbox{Graduation}_{i}) = \beta_{0}\mbox{Probation}_{i} + f(\mbox{Cutoff}\:\mbox{Distance}_{i}) + \alpha_{1}\mbox{Gender}_{i}+\alpha_{2}\mbox{HighSchool}_{i}+\alpha_{3}\mbox{Language}_{i} +\epsilon_{i}$$
&nbsp; &nbsp; The model baseline characteristics are the same as mentioned in dropout Section \@ref(baseline). As shown in the following table (Table \@ref(tab:gradsummary)), within 0.2 GPA bandwidth, compared to students on academic probation, students in good standing have $26.8\%$ increase in the odds of graduating by Year 4, while holding other baseline characteristics constant. **However**, the p-value of 0.161 suggests that the graduation does not significantly differ between good standing students and probation students, holding other baseline characteristics constant. And this difference could also not be clearly detected around the probation cutoff as shown (Figure \@ref(fig:gradfig)). Within 0.3 GPA and 0.6 GPA bandwidth, the results also do not suggest difference on graduation between good standing students and probation students at 0.05 significance level, holding other baseline characteristics constant.

@@

```{r gradfig, fig.align='center', fig.cap="Academic Probation Impact on Graduation by Year 4", fig.width=8, fig.height=4, echo= FALSE, message=FALSE, warning=FALSE}
clean_cut$probation_year1 <- as.factor(clean_cut$probation_year1)
ggplot(data= clean_cut, aes(x = dist_from_cut, y = gradin4, color = status))+
  geom_point(alpha = 0.05)+
  geom_smooth(method = "glm", method.args = list(family = 'binomial'))+
  theme_light()+
  geom_vline(xintercept = 0, linetype = "dotted")+
  theme(legend.title = element_blank(), legend.position = "bottom", plot.title = element_text(hjust = 0.5))+
  guides(color=guide_legend(override.aes=list(fill=NA)))+
  labs(title = "Academic Probation Impact on Graduation by Year 4",
       x = "Year 1 GPA minus Probation Cutoff",
       y = "Graduate by Year 4")
```

@@

&nbsp; &nbsp; Since all result interpretations are based on p-values, reassessing the definition of p-value may help to explain the above inconsistent results. In simple common English, P-value means the probability that something checked wrong to happen. The lower p-value means a higher probability of something checked right to be true, and "is there an academic probation impact on first year student outcome" is the checkpoint in this study.

@@

&nbsp; &nbsp; According to the student summary statistics (Table \@ref(tab:lindo)) and t-test conducted on students highschool performance (Table \@ref(tab:ttesthslin)) and credit attempted in the first year (Table \@ref(tab:ttestcrlin)) in Data Section \@ref(bandwidth), within the bandwidth of 0.6 GPA, students below the cutoff generally hold weaker academic potentials than those above the cutoff. And these academic differences also serve as key factors affecting student outcomes.

&nbsp; &nbsp; Even though Lindos's original paper did test whether students share similar observable characteristics around the probation cutoff GPA in Results Section A [@cite_original], it failed to access student characteristics that are further away from the cutoff. It is unlikely that a student with 1.0 GPA is as similar as one with 2.0 GPA, while they are all included within Lindo's bandwidth to estimate the impact of academic probation. As a result, using a wider bandwidth could violate the similarity assumption of RDD, and interpreting p-values from assumption violated models to seek "statistical significance" may become senseless.

@@

&nbsp; &nbsp; The Credit/No Credit option allows students to remove course result from their GPA but still count course credit if passing the course (at least $50\%$ final mark, 2.0 credits maximum) before a certain deadline (usually in the middle of the semester). For example, if a student utilizes this policy on:

@@

## Enhancement {#enhancement}
